---
title: "4-GBIF"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## GBIF Data
An idea on how to plot GBIF data.
https://www.gbif.org/

This is a way to make your own range maps. There are several other ways to go about retrieving data from GBIF such as the "rgbif" R package.

"bdvis" provides downloading/mapping instructions that may work for you:
https://doi.org/10.1093/bioinformatics/btw333

It didn't work for me, so I downloaded the GBIF data myself.

## Watch out for....

* Coordinates: Does the entry have them?
* Sampling bias? Some places have more records than others...
* What is classified. Pleuronectidae for example contains a few other families on GBIF. 
* How specific is the identification? To species? To genus?

Pleuronectiformes also creates a 357 Mb data file. Downloading via an API (with rgbif) is a bit cumbersome at that point.

## Starting data.
I have created a file for all Bothidae occurrence records with GBIF: https://doi.org/10.15468/dl.mpxaok

It is too large to synch well with GitHub... So, I made it smaller by only keeping a few key fields:

family	genus	species	decimalLatitude	decimalLongitude

This reduced file is in /4-GBIF/ folder and is named "bothidae.csv" and is ~ 3 Mb.

#### Side Note: How I reduced the file size with bash.
I found the GBIF file to be a problem to parse with all the fields as well making import in R pretty much impossible for me.

Here is how I reduced the file to columns 8,9,10,17, and 18 at the command line. This is not with R.

```{r, engine = 'bash', eval = FALSE}
cat 0008756-181108115102211.csv | cut -f 8,9,10,17,18 > bothidae.csv 
```



## Starting out
What does the data file look like? What do we want in the end to make a raster? You can open it with a text editor and take a look. 

What we need to do do is import the data into R. Like last time, I made a tibble.

* I used the tidyverse library (and dplyr)
* To load the file, I used read.table
* It is important to consider missing entries in the table with GBIF data...
* I used as_tibble
* I used rename

Create a tibble like so:

```{r, echo=FALSE, warnings=FALSE, message=FALSE}
library("tidyverse")
gbif<-read.table(file="bothidae.csv", header=TRUE, sep="\t",
                 fill = TRUE) 

gbifdf<-as_tibble(gbif) %>%
 rename(Latitude=decimalLatitude, Longitude=decimalLongitude) %>%
  rename(Family=family, Genus=genus, Species=species)

gbifdf
```

## What do we want again?
We would like to have an estimate of number of species within an a degree grid square (e.g. species in 1 deg Latitude by 1 deg Longitude). You may have noticed that we have entries just for Bothidae with no species. Let's remove those.

* Use dplyr filter and a Boolean operator
```{r,  warnings=FALSE, message=FALSE, echo=FALSE}
df<-gbifdf %>% filter(Species != "")

df
```

Should be ~7,000 less entries.

## How to get species diversity within a grid square...

### Creating an using a function in R.
Converting our point collections to species is complicated. I have made a function and stored it in an external file, rasterize.R. This is the basics of using a function in R.

The basic form of a function in R:

* Declare variables e.g. (x,y)
* Do something to variables (x*y)
* Return something else

```{r}
Multiply <- function(x,y) {
product<-x*y
return(product)
}
```

Here is how to use the function:
```{r}
Multiply(3,5)
```

The function should show up in your Environment tab within RStudion.

### Rasterize
The Rasterize function I wrote can be loaded with source. Can you source it?
```{r}
source("rasterize.R")
```
Rasterize requires three things:

1. cellSize -> the grid square ofLongitude/Latidue as a number, e.g. 2
2. breaks -> A vector to combine species counts e.g. c("10", "20", "30")
3. dataFrame -> A data frame with Family, Genus, Species, Latitude, and Longitude columns.

Using it:
```{r}
cellSize<-3
breaks<-seq(0, length(levels(df$Species)), 5) # From 0 to 160 here by 10
                   
raster<-Rasterize(cellSize, breaks, df)
raster
```

## Plotting diversity within grid squares.

There are two options here. The Rasterize funciton returns a strict diversity count (1,2,3,4,5) or a range 1-10, 11-20, 21-30, etc.

#### Plotting with "Diversity":

```{r}
ggplot()+geom_raster(data=raster, aes(x=Longitude,y=Latitude, 
                                      fill=Diversity))+
  coord_fixed(1.3)+
  theme_classic()
```

* Can you use "scale_fill_continuous" to change the color scheme?

```{r, echo=FALSE}
ggplot()+geom_raster(data=raster, aes(x=Longitude,y=Latitude, 
                                      fill=Diversity))+
  scale_fill_continuous(low="yellow", high="red")+
  coord_fixed(1.3)+
  theme_classic()
```

#### Plotting with "Breaks"
Plotting with "Breaks" requires us to set our own color scale. I used the "heat.colors" function.

```{r}

heatColors<-heat.colors(10)

ggplot()+geom_point(aes(x=seq(1,10,1), y=0), cex=10, color=heatColors)

```

```{r, eval=FALSE}
number<-length(unique(raster$Breaks)) # Number of colors we need

colors<-rev(heat.colors(number)) #heat.colors produces the number of colors, with yellow high and red low, so I reversed it.
```

We can incldue this custom scale in ggplot with the following command:
```{r, eval=FALSE}
scale_fill_manual(values=colors)
```

```{r, echo=FALSE}
number<-length(unique(raster$Breaks))
colors<-rev(heat.colors(number))

ggplot()+geom_raster(data=raster, aes(x=Longitude,y=Latitude, 
                                      fill=Breaks))+
  scale_fill_manual(values=colors)+
  coord_fixed(1.3)+
  theme_classic()
```

## What do you want to plot?
Why don't you download your own taxon of interest and try to plot?

```{r, echo=FALSE, warnings=FALSE, message=FALSE}
library(maps)
library(mapdata)

world<-map_data("world")
#Create to identical objects from world
w1 <- world
w2 <- world
# Make one go 360 degrees longer
w2$long <- w2$long + 360
# Extend the groups so that they are not redundant
w2$group <- w2$group + max(w2$group) + 1
#Combine them
newWorld <- rbind(w1, w2)
newWorld <- newWorld %>% rename(Latitude=lat, Longitude=long)

# doubling the raster
raster2<-raster
raster2$Longitude <- raster2$Longitude+360


map<-ggplot(newWorld)+geom_polygon(aes(x=Longitude, y=Latitude,
                                  group=group),  color="black",
                                  fill="black", size=0.5)+
  geom_raster(data=raster, aes(x=Longitude,y=Latitude, 
                                      fill=Breaks))+
  geom_raster(data=raster2, aes(x=Longitude,y=Latitude, 
                                      fill=Breaks))+
  scale_fill_manual(values=colors)+
  coord_fixed(ratio=1.3, xlim = c(-0, 360), ylim=c(-90,90))+
  theme_classic()+
  theme(panel.background = element_rect(fill="lightblue"))

map
```